Train Harder, Smarter:\\Using Graph Theory to Create Route Suggestions

### 2013-11-19 paper.tex
Added a few images to the writeup.

\subsection{Distance Calculations}
As mentioned earlier, using the Euclidian norm for latitudes and longitudes
is highly inaccurate; latitudes and longitudes are more similar to 3D polar
coordinates with a fixed $r$ than they are to a coordinate plane. The
haversine formula is used to calculate distances between two points on the
surface of a sphere. The haversine formula uses the haversin function,
which is defined as follows:
\newcommand{\haversin}{\mbox{haversin}}
\[ \haversin (\theta) = \sin^2(\frac{\theta}{2}). \]
Given two coordinate pairs, $(\phi_1, \lambda_1)$ and $(\phi_2, \lambda_2)$,
and a radius, $r$, the haversine formula calculates the distance $d$ between
the two coordinate pairs:
\begin{align*}
  d = r\haversin^{-1}(&\haversin(\phi_2 - \phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2 -\lambda_1)
\end{align*}
Using the definition of the haversin function, this can be reduced into a
function in terms of elementary trigonometric functions. The only non-trivial
thing to note is that $\haversin^{-1}(x) = 2\sin^{-1}(\sqrt{x})$. Armed with
this knowledge, it is fairly easy to reduce. We'll use an intermediate variable
to hold the value under the square root:
\begin{align*}
  h = & \haversin(\phi_2-\phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2-\lambda_1) \\
  = & \sin^2\left(\frac{\phi_2 - \phi_1}{2}\right) + \\
  & \cos(\phi_1)\cos(\phi_2)\sin^2\left(\frac{\lambda_2-\lambda_1}{2}\right)\\
  d = & r\haversin^{-1}\left(\sqrt{h}\right) \\
  = & 2r\sin^{-1}\left(\sqrt{h}\right)
\end{align*}
By simply substituting this as the right hand side of \eqref{eq:dj}, many of the
projection issues can be avoided.

@@

To find the proper loop, a cone must be projected from the origin of the sphere,
and the intersection of the cone and the sphere can be used to calculate the
ideal loop. The following equation describes a cone,
\[ 0 = (u \cdot d)^2 - |d||u|\cos \phi , \]
where $u= (x,y,z)$, $d = (a,b,c)$ is a vector parallel to the axis of the
cone, and $\phi$ is one-half the aperture of the cone. Additionally, recall
the equation for a sphere:
\[ r = |u|. \]
Unfortunately, both of these equations are of little aid, since they require
input in the form of cartesian coordinates. We would like to convert these to
a set of equations in spherical form. However, these equations become rather
complicated; for now, we've accepted the projected sphere as ``good enough.''

>>>>

As mentioned earlier, using the Euclidian norm for latitudes and longitudes
is highly inaccurate; latitudes and longitudes are spherical coordinates with
fixed $r$, so considering the latitude and longitude to be $x$ and $y$ values is
a very flawed idea. As one can see in Figure~\ref{fig:sphere}, the contrast is
stark. Latitude and longitudes are angles, not positions in space.

\begin{figure}
  \includegraphics[width=3in]{images/558px-3D_Spherical.png}
  \caption{The Spherical Coordinate System} \label{fig:sphere}
\end{figure}

\subsection{Distance Calculations}
The haversine formula is used to calculate distances between two points on the
surface of a sphere. The haversine formula uses the haversin function,
which is defined as follows:
\newcommand{\haversin}{\mbox{haversin}}
\[ \haversin (\theta) = \sin^2(\frac{\theta}{2}). \]
Given two coordinate pairs, $(\phi_1, \lambda_1)$ and $(\phi_2, \lambda_2)$,
and a radius, $r$, the haversine formula calculates the distance $d$ between
the two coordinate pairs:
\begin{align*}
  d = r\haversin^{-1}(&\haversin(\phi_2 - \phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2 -\lambda_1)
\end{align*}
Using the definition of the haversin function, this can be reduced into a
function in terms of elementary trigonometric functions. The only non-trivial
thing to note is that $\haversin^{-1}(x) = 2\sin^{-1}(\sqrt{x})$. Armed with
this knowledge, it is fairly easy to reduce. We'll use an intermediate variable
to hold the value under the square root:
\begin{align*}
  h = & \haversin(\phi_2-\phi_1) + \\
  & \cos(\phi_1)\cos(\phi_2)\haversin(\lambda_2-\lambda_1) \\
  = & \sin^2\left(\frac{\phi_2 - \phi_1}{2}\right) + \\
  & \cos(\phi_1)\cos(\phi_2)\sin^2\left(\frac{\lambda_2-\lambda_1}{2}\right)\\
  d = & r\haversin^{-1}\left(\sqrt{h}\right) \\
  = & 2r\sin^{-1}\left(\sqrt{h}\right)
\end{align*}
By simply substituting this as the right hand side of \eqref{eq:dj}, many of the
projection issues can be avoided.

@@

\begin{figure}
  \centering
  \includegraphics[width=2in]{images/220px-Numerical_aperture.png}
  \caption{The aperture of a cone} \label{fig:aperture}
\end{figure}
To find the proper loop, a cone must be projected from the origin of the sphere,
and the intersection of the cone and the sphere can be used to calculate the
ideal loop. The following equation describes a cone,
\[ 0 = (u \cdot d)^2 - |d||u|\cos \phi , \]
where $u= (x,y,z)$, $d = (a,b,c)$ is a vector parallel to the axis of the
cone, and $\phi$ is one-half the aperture of the cone. In
Figure~\ref{fig:aperture}, the $\theta$ pictured is the same as the $\phi$ in
this equation. Additionally, recall
the equation for a sphere:
\[ r = |u|. \]
For the most part, we can consider $r = 1$, since we are dealing with the
surface of a sphere. However, we do have to make certain calculations using
the actual radius of the earth. For example, in order to calculate the
aperture $\phi$, we realize that the length of the sides of the cone will be the
radius of the earth. Then $\sin \phi = \frac{r_c}{r}$, so we can calculate
$\phi$ as follows:
\begin{equation}
  \phi = \sin^{-1} \left( \frac{\frac{d}{2\pi}}{r} \right) \label{eq:phi}
\end{equation}

Unfortunately, the equations for spheres and cones are of little aid, since
they require
input in the form of cartesian coordinates. We would like to convert these to
a set of equations in spherical form. However, these equations become rather
complicated; for now, we've accepted the projected circle as ``good enough.''
